# Examining data I

## Introduction {#introduction-w02}
Welcome to your second week of **Introduction to Quantitative Research Methods**. This week we will focus on **examining data** using measures of central tendency and measures of dispersion. These measures are collectively known as **descriptive statistics**. We will also talk about some basic data visualisation. Also, as by now, everyone should be up and running with RStudio Server, we will apply some of these descriptive measures onto some data. Alright, let's get to it.

### Reading list

Please find the reading list for this week below. We strongly recommend that you read the core reading material before you continue with the rest of this week’s material.

- Cetinkaya-Rundel, M., and Hardin, J., 2021, **Introduction to Modern Statistics**, Section II: Exploratory data analysis, Chapter 5: Exploring numerical data, Page(s): 76-95. Source: [**openintro.org/book/ims**](https://openintro.org/book/ims).

To download the full book, click on this link: [**https://openintro-ims.netlify.app**](https://openintro-ims.netlify.app), Then make sure to click following: "**Download PDF**" >> "**Read Free Sample**" >> "**PDF**"   

### Q & A session

This week there will be a live Q&A session on **Tuesday, January 18, 2022 @ 11h00 to 12:30 (GMT)**. The session will be hosted on Zoom [**LINK**](https://ucl.zoom.us/j/96460973152?pwd=K2taYWJLdm5HKzAzQXJCUThxbUczUT09). For security reasons, the meeting password for this Q&A session can be found on Moodle on the main page: **POLS0008 Online Q&A - Tuesdays 11:00-12:30**

## Measures of central tendency
Any research project involving [quantitative data](https://en.wikipedia.org/wiki/Quantitative_research) should start with an exploration and examination of the available data sets. This applies both to data that you have collected yourself and data that you have acquired in a different way, e.g., through downloading official [UK Census and labour market statistics](https://www.nomisweb.co.uk/). The set of techniques that is used to examine your data in first instance is called [descriptive statistics](https://en.wikipedia.org/wiki/Descriptive_statistics). Descriptive statistics are used to describe the basic features of your data set and provide simple summaries about your data. Together with simple visual analysis, they form the basis of virtually **every** quantitative data analysis.

### Accessing RStudio Server, uploading and importing data

For this tutorial in week 2, we will use the `London historical population dataset.csv`. You can download the dataset from [**Moodle**](https://moodle.ucl.ac.uk/mod/resource/view.php?id=3326334). Let's us first:

- Log into the RStudio Server: [**https://rstudio.data-science.rc.ucl.ac.uk/**](https://rstudio.data-science.rc.ucl.ac.uk/). Log in with your usual UCL **username** and **password**.
- Create a new folder called `Week 2`
- Upload the data `London historical population dataset.csv` into the folder of `Week 2`
- Set the directory to the folder location of `Week 2` to where the CSV file was uploaded by clicking on the `More` >> `Set As Working Directory`.
- Finally, import the data into workspace using the `read.csv()` function.

```{r}
# Load data into RStudio. The spreadsheet is stored in the object called 'London.Pop'
London.Pop <- read.csv("London historical population dataset.csv")
```

If you struggle with the above steps, as well as setting up your working directory. Then here is a reminder, [have a look at how we did this last week](https://uclpg-msc-sgds.github.io/POLS0008/understanding-data.html#importing-data-into-rstudio)!

Use the `head()`, or `View()` functions to examine the structure of the London population data frame. The `head()` functions allows the users to see the first 5 rows of the data frame in the R-Console. The View() allows the user to see the entire spreadsheet in a data viewer.

### Mean, Median and Mode

We are going to use the following functions: `mean()` and `median()`, to compute the **mean** and **median** value respectively - these two estimates are one of the three **measures of central tendency**. 

We can apply these measures on our `London.Pop` data set. Let's do this by focusing on London's population in 2011. We going to keep the 1st, 2nd and 24th columns in the `London.Pop` data frame which corresponds to the columns called `Area Code`, `Area Name` and `Person-2011`, respectively.

```{r, warnings=FALSE, message=FALSE}
# select the relevant columns
London.Pop2011 <- London.Pop[,c(1:2,24)]
```

Now we can calculate our **measures of central tendency** using R's built-in functions for the median and the mean.

```{r, warnings=FALSE, message=FALSE }
# calculate the median of the 2011 population variable
median(London.Pop2011$Persons.2011)
# calculate the mean of the 2011 population variable
mean(London.Pop2011$Persons.2011)
```

#### Question(s) {-}
1. How do you explain that the median is larger than the mean for this variable? [**HINT**: What is distorting the mean value such that its brought the value down?]

R does not have a standard in-built function to calculate the mode. As we still want to show the mode, we create a user function to calculate the mode of our data set. This function takes [a numeric vector](https://r4ds.had.co.nz/vectors.html) as input and gives the mode value as output.

<div class="note">
**Note**<br/>
You do not have to worry about creating your own functions, so just copy and paste the code below to create the `get_mode()` function. 
</div>

```{r, warnings=FALSE, message=FALSE }
# create a function to calculate the mode
get_mode <- function(x) {
  
  # get unique values of the input vector
  uniqv <- unique(x)
  
  # select the values with the highest number of occurrences
  uniqv[which.max(tabulate(match(x, uniqv)))]
}
# calculate the mode of the 2011 population variable
get_mode(London.Pop2011$Persons.2011)
```

#### Question(s) {-}
1. What is the level of measurement of our *Persons.2011* variable? Nominal, ordinal, or interval/ratio?
2. Even though we went through all the trouble to create our own function to calculate the mode, do you think it is a good choice to calculate the mode for this variable? Why? Why not?

Although R does most of the hard work for us, especially with the `mean()` and the `median()` function, it is a good idea to once go through the calculations of these two central tendency measures ourselves. Let's calculate the mean step-by-step and then verify our results with the results of R's `mean()` function.

```{r, warnings=FALSE, message=FALSE }
# get the sum of all values
Persons.2011.Sum <- sum(London.Pop2011$Persons.2011)
# inspect the result
Persons.2011.Sum
# get the total number of observations
Persons.2011.Obs <- length(London.Pop2011$Persons.2011)
# inspect the result 
Persons.2011.Obs
# calculate the mean
Persons.2011.Mean <- Persons.2011.Sum / Persons.2011.Obs
# inspect the result
Persons.2011.Mean
# compare our result with R's built-in function
mean(London.Pop2011$Persons.2011) == Persons.2011.Mean
```

Great. Our own calculation of the mean is identical to R's built-in function. Now let's do the same for the median.

```{r, warnings=FALSE, message=FALSE }
# get the total number of observations
Persons.2011.Obs <- length(London.Pop2011$Persons.2011)
# inspect the result
Persons.2011.Obs
# order our data from lowest to highest
Persons.2011.Ordered <- sort(London.Pop2011$Persons.2011, decreasing=FALSE)
# inspect the result
Persons.2011.Ordered
# get the number of the observation that contains the median value
Persons.Median.Obs <- (Persons.2011.Obs + 1)/2
# inspect the result
Persons.Median.Obs
# get the median
Persons.2011.Median <- Persons.2011.Ordered[Persons.Median.Obs]
# inspect the result
Persons.2011.Median
# compare our result with R's built-in function
median(London.Pop2011$Persons.2011) == Persons.2011.Median
```

## Simple plots
Before moving on to the second set of descriptive statistics, the **measures of dispersion**, this is a good moment to note that simple data visualisations are also an extremely powerful tool to explore your data. In fact, tools to create high quality plots have become one of R's greatest assets. This is a relatively recent development since the software has traditionally been [focused on the statistics](https://en.wikipedia.org/wiki/R_(programming_language)) rather than visualisation. The standard installation of R has base graphic functionality built in to produce very simple plots. For example we can plot the relationship between the London population in 1811 and 1911. 

<div class="note">
**Note**<br/>
Next week we will be diving deeper into data visualisation and making plots and graphs in R, but for now it is a good idea to already take a sneak peek at how to create some basic plots.
</div>

```{r, warnings=FALSE, message=FALSE}
# make a quick plot of two variables of the London population data set
plot(London.Pop$Persons.1811,London.Pop$Persons.1911)
```

#### Experimenting with `plot()` {-}
1. What happens if you change the order of the variables you put in the `plot()` function? Why?
2. Instead of using the `$` to select the columns of our data set, how else can we get the same results?

The result of calling the `plot()` function, is a very simple [scatter graph](https://en.wikipedia.org/wiki/Scatter_plot). The `plot()` function offers a huge number of options for customisation. You can see them using the `?plot` help pages and also the `?par` help pages (`par` in this case is short for parameters). There are some examples below (note how the parameters come after specifying the `x` and `y` columns).

```{r, warnings=FALSE, message=FALSE}
# add a title, change point colour, change point size
plot(London.Pop$Persons.1811, London.Pop$Persons.1911, main='Quick Plot in R', col='blue', cex=2)
```

```{r, warnings=FALSE, message=FALSE}
# add a title, change point colour, change point symbol
plot(London.Pop$Persons.1811, London.Pop$Persons.1911, main="Another Quick Plot in R", col='magenta', pch=22)
```

Also, you can apply titles to the axes using the `xlab =""` and `ylab=""` argument after the `main=""` in the `plot()`. 

```{r, warnings=FALSE, message=FALSE}
# add a axis titles, rotate number labels
plot(London.Pop$Persons.1811, London.Pop$Persons.1911, main="Another Quick Plot in R", xlab = "Population in 1811", ylab = "Population in 1911", col='magenta', pch=22)
```

You can prevent R from printing the scientific notation (e.g., `2e+05`, `4e+05` etc.,) on the y-axis in the graphical output. You can turn it off by typing the following code before running the `plot()` function

```{r, eval=FALSE}
# turn-off horrible scientific notation
options(scipen = 999)
# add a axis titles, rotate number labels
plot(London.Pop$Persons.1811, London.Pop$Persons.1911, main="Another Quick Plot in R", xlab = "Population in 1811", ylab = "Population in 1911", col='magenta', pch=22)
```

<div class="note">
**Note**<br/>
For more information on the plot parameters (some have obscure names) have a look here: http://www.statmethods.net/advgraphs/parameters.html
</div>

#### Recap {-}
In this section you have learnt how to:

1. Calculate the mode, median, and mean of a variable in R.
2. Make some simple scatter plots (in preparation for next week) to visualise your data.

## Measures of dispersion
When exploring your data, **measures of central tendency** alone are not enough as they only tell you what a 'typical' value looks like but they do not tell you anything about all other values. Therefore we also need to look at some **measures of dispersion**. Measures of dispersion describe the spread of data around a central value (e.g. the mean, the median, or the mode). The most commonly used measure of dispersion is the [standard deviation](https://en.wikipedia.org/wiki/Standard_deviation). The standard deviation is a measure to summarise the spread of your data around the mean. The short video below will introduce you to the standard deviation as well as to three other measures of dispersion: the **range**, the **interquartile range**, and the **variance**.

For the rest of this tutorial we will change our data set to one containing the number of assault incidents that ambulances have been called to in London between 2009 and 2011. You will need to download a prepared version of this file called: `Ambulance and Assault Incidents data.csv` and upload it to your working directory. It is in the same data format (`csv`) as our London population file so we use the `read.csv()` command again.

```{r}
# Load data into RStudio. The spreadsheet is stored in the object called 'London.Ambulance'
London.Ambulance <- read.csv('Ambulance and Assault Incidents data.csv')
```

```{r, warnings=FALSE, message=FALSE}
# inspect the results 
head(London.Ambulance)
```

```{r, warnings=FALSE, message=FALSE}
# inspect the number of rows and columns of the data set
dim(London.Ambulance)
```

You will notice that the data table has 5 columns and 649 rows. The column headings are abbreviations of the following:

| Column heading | Full name | Description |
| :---           | :---      | :---------- |
|**BorCode** | Borough Code | London has 32 Boroughs (such as Camden, Islington, Westminster, etc.) plus the City of London at the centre. These codes are used as a quick way of referring to them from official data sources. 
| **WardName** | Ward Name | Boroughs can be broken into much smaller areas known as [Wards](https://en.wikipedia.org/wiki/Wards_and_electoral_divisions_of_the_United_Kingdom). These are electoral districts and have existed in London for centuries. | 
| **WardCode** | Ward Code | A statistical code for the wards above. |
| **WardType** | Ward Type | A classification that groups wards based on similar characteristics. |
| **Assault_09_11** | Assault Incidents | The number of assault incidents requiring an ambulance between 2009 and 2011 for each ward in London. |

Let's start by calculating two measures of central tendency by using the `median()` and `mean()` functions.

```{r, warnings=FALSE, message=FALSE}
# calculate the median of the assault incident variable
median(London.Ambulance$Assault_09_11)
# calculate the mean of the assault incident variable
mean(London.Ambulance$Assault_09_11)
```

#### Questions {-}
1. How do you explain that the mean is larger than the median for this variable?

Great. Let's now calculate some measures of dispersion for our data: the range, the interquartile range, and the standard deviation. The calculation of the range is very straightforward as we only need to subtract the the minimum value from the the maximum value. We can find these values by using the built-in `min()` and `max()` functions.

```{r 02-calculate-that-range, warnings=FALSE, message=FALSE}
# get the minimum value of the assault incident variable
min(London.Ambulance$Assault_09_11)
# get the maximum value of the assault incident variable
max(London.Ambulance$Assault_09_11)
# calculate the range
1582 - 0
# or in one go
max(London.Ambulance$Assault_09_11) - min(London.Ambulance$Assault_09_11)
```

#### Questions {-}
1. What does this range mean? 

The interquartile range requires a little bit more work to be done as we now need to work out the values of the 25th and 75th [percentile](https://en.wikipedia.org/wiki/Percentile).

<div class="note">
**Note**<br/>
A percentile is a score at or below which a given percentage of your data points fall. For example, the 50th percentile (also known as the median!) is the score at or below which 50% of the scores in the distribution may be found.
</div>

```{r, warnings=FALSE, message=FALSE}
# get the total number of observations
London.Ambulance.Obs <- length(London.Ambulance$Assault_09_11)
# inspect the result
London.Ambulance.Obs
# order our data from lowest to highest
London.Ambulance.Ordered <- sort(London.Ambulance$Assault_09_11, decreasing=FALSE)
# inspect the result
London.Ambulance.Ordered
# get the number ('index value') of the observation that contains the 25th percentile
London.Ambulance.Q1 <- (London.Ambulance.Obs + 1)/4
# inspect the result
London.Ambulance.Q1
# get the number ('index value') of the observation that contains the 75th percentile
London.Ambulance.Q3 <- 3*(London.Ambulance.Obs + 1)/4
# inspect the result
London.Ambulance.Q3
# get the 25th percentile
London.Ambulance.Ordered[163]
# get the 75the percentile
London.Ambulance.Ordered[488]
# get the interquartile range
233 - 86
```

As explained at the very end of this [**lecture video**](https://web.microsoftstream.com/video/5d95c429-4c0d-451f-8359-2cb9729eb837), we can also visually represent our range, median, and interquartile range using a box and whisker plot:

```{r, warnings=FALSE, message=FALSE}
# make a quick boxplot of our assault incident variable
boxplot(London.Ambulance$Assault_09_11, horizontal=TRUE)
```

#### Questions {-}
1. There is a large difference between the range that we calculated and the interquartile range that we calculated. What does this mean?
2. The 25th and 75th percentile in the example do not return integer but a fraction (i.e. 162.5 and 487.5). Why do we use 163 and 488 to extract our percentile values and not 162 and 487?

Now, let's move to the standard deviation. Remember: this is one of the most important measures of dispersion and is widely used in all kinds of statistics. The calculation involves the following steps as extensively explained in this [**lecture video**](https://web.microsoftstream.com/video/8f043b29-3e71-4ef7-85f2-6c7a6dd1f090):

1. Calculate the mean.
2. Subtract the mean from each observation to get a residual.
3. Square each residual.
4. Sum all residuals.
5. Divide by $n-1$.
6. Take the square root of the final number.

```{r, warnings=FALSE, message=FALSE}
# calculate the mean
London.Ambulance.Mean <- mean(London.Ambulance$Assault_09_11)
# subtract the mean from each observation
London.Ambulance.Res <- London.Ambulance$Assault_09_11 - London.Ambulance.Mean
# square each residual
London.Ambulance.Res.Sq <- London.Ambulance.Res**2
# sum all squared residuals
London.Ambulance.Res.Sum <- sum(London.Ambulance.Res.Sq)
# divide the sum of all sqaured residuals by n - 1
London.Ambulance.Variance <- London.Ambulance.Res.Sum / (length(London.Ambulance$Assault_09_11) - 1)
# take the square root of the final number
London.Ambulance.Sd <- sqrt(London.Ambulance.Variance)
# standard deviation
London.Ambulance.Sd
```

There we go. We got our standard deviation! You probably already saw this coming, but R does have some built-in functions to actually calculate these descriptive statistics for us: `range()`, `IQR()`, and `sd()` will do all the hard work for us!

```{r, warnings=FALSE, message=FALSE}
# range
range(London.Ambulance$Assault_09_11)
# interquartile range
IQR(London.Ambulance$Assault_09_11)
# standard deviation
sd(London.Ambulance$Assault_09_11)
```

<div class="note">
**Note**<br/>
Please be aware that the `IQR()` function may give slighlty different results in some cases when compared to a manual calculation. This is because the forumula that the `IQR()` function uses is slightly different than the formula that we have used in our manual calculation. It is noted in the documentation of the `IQR()` function that: "Note that this function computes the quartiles using the quantile function rather than following Tukey's recommendations, i.e., `IQR(x) = quantile(x, 3/4) - quantile(x, 1/4)`."
</div>

#### Questions {-}
1. What does it mean that we have a standard deviation of 130.3482?
2. Given the context of the data, do you think this is a low or a high standard deviation?

To make things even easier, R also has a `summary()` function that calculates a number of these routine statistics simultaneously. After running the `summary()` function on our assault incident variable, you should see you get the minimum (Min.) and maximum (Max.) values of the *assault_09_11* column; its first (1st Qu.) and third (3rd Qu.) quartiles that comprise the interquartile range; its the mean and the median. The built-in R `summary()` function does not calculate the standard deviation. There are functions in other libraries that calculate more detailed descriptive statistics, 

```{r, warnings=FALSE, message=FALSE}
# calculate the most common descriptive statistics for the assault incident variable
summary(London.Ambulance$Assault_09_11)
```

#### Recap {-}
In this section you have learnt how to:

1. Calculate the range, interquartile range, and standard deviation in R.
2. Create a simple boxplot using the `boxplot()` function.
3. Quickly get common descriptive statistics using the `summary()` function.

## Seminar {#seminar-w02}

Please find the seminar task and seminar questions for this week's seminar below. 

<div class="note">
**Note**<br/>
Please make sure that you have executed the seminar task and have answered the seminar questions before the seminar!
</div>

#### Seminar task {-}
Using the `London.Ambulance` object:

1. Create a new object / data set that **only** contains data for ward type *Multicultural Metropolitan*. 
2. Create a new object / data set that **only** contains data for ward type *Prospering Metropolitan*.

<div class="note">
**Note**<br/>
Creating a new object from an existing object is called creating a **subset** of your data. There are many ways of creating a subset, but one way of doing it is by using the following syntax: `London.Assault[London.Assault$WardType == 'Multicultural Metropolitan',]`.
</div>

For **both** new objects:

1. Calculate the median, mean, range, interquartile range, and standard deviation for the *Assault_09_11* variable.
2. Create a boxplot of the *Assault_09_11* variable.

#### Seminar questions {-}

1. Explain why each of these statistics are useful and what type of data are required to calculate them:
- Mode
- Median
- Mean
- Range
- Interquartile Range
- Standard Deviation
2. Compare the results of the descriptive statistics you have calculated for your *Multicultural Metropolitan* object / data set with the results of the descriptive statistics you have calculated for you *Prospering Metropolitan* object / data set. What do these differences tell us? In which of these two ward types are the data more clustered? In which of these two ward types are the data more dispersed? 

#### Seminar link {-}
Seminars for all groups take place on **Thursday morning**. You can find the **Zoom** link to your seminar group on [Moodle](https://moodle.ucl.ac.uk/course/view.php?id=6845). 

## Before you leave {#byl-w02}
Save your R script by pressing the **Save** button in the script window. [That is it for this week](https://www.youtube.com/watch?v=Wmc8bQoL-J0)!